---
title: "SLHD-day1"
author: "CB"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The learning process for supervised learning

In order to illustrate the evaluation of the performance of supervised classifiers, let's first consider some toy data: the Fisher's Iris

```{r}
data(iris)
?iris
```

```{r}
dim(iris)
```

```{r}
colnames(iris)
```

```{r}
X = iris[,-5]
Y = iris[,5]
```

> Exercise: have a look at the documentation of LDA (`lda` function in the package `MASS`) to learn a LDA classifier on the iris data.

```{r}
library(MASS)
?lda
```

```{r}
f = lda(X,Y)
f
```



Then, we can use this classifier to predict the class of a specific data point `x.star = c(6,2.8,5,1.6)`

> Exercise: search in the documentation of lda how to use the learned classifier to predict the class of a new data point.

```{r}
x.star = c(6,2.8,5,1.6)
predict(f,x.star)
```

> Exercise: evaluate the learning classification error (the error which is too optimistic, the one not to consider) on this data set.

```{r}
class = predict(f,X)$class
err.learn = sum(class != Y) / length(Y)
err.learn
```

> Exercise: apply the minimal setup to evaluate in a better manner the classifcation error (in validation) of LDA on this data set. As the data are sorted here by class, we first need to shuffle the data. 

```{r}
# Shuffling the data
ord = sample(nrow(X),nrow(X))
X = X[ord,]
Y = Y[ord]

# Computation of the validation error
X.learn = X[1:100,]
Y.learn = Y[1:100]
X.val = X[101:150,]
Y.val = Y[101:150]

f = lda(X.learn,Y.learn)
class =  predict(f,X.val)$class
err.val = sum(class != Y.val) / length(Y.val)
err.val
```

In order to have a better evaluation than the one of the minimal setup, we need to code a V-fold cross-validation:

```{r}
V = 15
folds = rep(1:V,nrow(X)/V)

err = rep(NA,V) 
for (v in 1:V){
  X.learn = X[folds!=v,]; Y.learn = Y[folds!=v]
  X.val = X[folds==v,]; Y.val = Y[folds==v]
  f = lda(X.learn,Y.learn)
  class =  predict(f,X.val)$class
  err[v] = sum(class != Y.val) / length(Y.val)
}
cat('LDA cv. error:',mean(err),'+/-',sd(err),'\n')
```

Let's now also consider the method k-NN, which is implemented in R in the `knn` function of the `class` package

```{r}
library(class)
?knn
```

> Exercise: learn a k-NN classifier with k=3 on the iris data and make a prediction for the class of `x.star = c(6,2.8,5,1.6)`

```{r}
x.star = c(6,2.8,5,1.6)
knn(X,x.star,Y,k = 3)
```

It is also interesting to see that changing the value of k actually changes the algorithm (and its prediction)

```{r}
knn(X,x.star,Y,k = 1)
knn(X,x.star,Y,k = 5)
```
we don't have the information about the error of classification as we do with LDA. 


This is a nice illustration that 1-NN and 5-NN are two different algorithms.

# Exercise: apply the minimum setup to evaluate the validation error of k-NN with k = 3.
```{r}

class =  knn(X.learn,X.val,Y.learn,k = 3)
err.val = sum(class != Y.val) / length(Y.val)
err.val
```








