---
title: "SLHD-day3"
author: "CB"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Support Vector Machines (2nd part)

In order to better explore the capacities of the different kernels in SVM, let's consider now the 2-dimensional 3-class classification problem of the iris data:


```{r}
library(MASS)
data(iris)
X = iris[,1:2]
Y = iris[,5]
plot(X,col=c('lightblue','pink','palegreen')[Y],pch=19)
```

Let's now compare the classification boundaries of the linear, polynomial (degree 2) and RBF kernels:

> Exercise: plot the classification boundaries of the linear kernel.

```{r}
library(e1071)
# Creating the grid of points to classify
xx = seq(min(X[,1])-0.1,max(X[,1])+0.1,length.out=200)
yy = seq(min(X[,2])-0.1,max(X[,2])+0.1,length.out=200)
Z = expand.grid(xx,yy)

# Learn the classifier in a traditional manner
# and classify all data points of the grid
f = svm(X,Y,type = "C",kernel = "linear")
class = predict(f,Z)

# Plot the classification boundaries
plot(Z,col=c('lightblue','pink','palegreen')[class],pch=19)
points(X,col=c('darkblue','red','darkgreen')[Y],pch=19)
```
We can see that with linear kernel, we have linear descision boudary.
> Exercise: modify this code to visualize the effect of varying the degree of the polynomial kernel on the classification boundaries.

```{r}
par(mfrow = c(2,3))

for (d in 1:6){
  f = svm(X,Y,type = "C",kernel = "poly",degree = d)
  class = predict(f,Z)
  
  # Plot the classification boundaries
  plot(Z,col=c('lightblue','pink','palegreen')[class],pch=19)
  points(X,col=c('darkblue','red','darkgreen')[Y],pch=19)
}
```

We can notice that odd degrees lead to less smooth classification boundaries and we may expect more validation error in these cases.

> Exercise: use CV to compare the classification errors of the different degrees.

```{r}
svm.bestD <- function(X,Y,V=15,d.max=15){
  folds = rep(1:V,nrow(X)/V)
  err = matrix(NA,V,d.max) 
  for (v in 1:V){
    X.learn = X[folds!=v,]; Y.learn = Y[folds!=v]
    X.val = X[folds==v,]; Y.val = Y[folds==v]
    for (d in 1:d.max){
      f = svm(X.learn,Y.learn,type="C",
              kernel="poly",degree = d)
      class = predict(f,X.val)
      err[v,d] = sum(class != Y.val) / length(Y.val)
    }
  }
  plot(colMeans(err),type='b')
  return(which.min(colMeans(err))) # Return the best d
}

d.star = svm.bestD(X,Y)
d.star
```

And now, let's consider the RBF kernel. First of all, let's have a look at the classification boundaries for different values of $\gamma$ (usually between $0.001$ and ... $10$).

```{r}
par(mfrow = c(2,3))

gamma = c(0.01,0.1,1,2,5,10)
for (g in gamma){
  f = svm(X,Y,type = "C",kernel = "radial",gamma = g)
  class = predict(f,Z)
  
  # Plot the classification boundaries
  plot(Z,col=c('lightblue','pink','palegreen')[class],pch=19)
  points(X,col=c('darkblue','red','darkgreen')[Y],pch=19)
}
```

> Exercise: Adapt the code above to be able to select the most appropriate value of the tuning parameter $\gamma$ using CV.

```{r}
svm.bestGamma <- function(X,Y,V=15,
        gamma = c(0.001,0.005,0.01,0.05,0.1,1,2,5,10,50,100)){
  folds = rep(1:V,nrow(X)/V)
  err = matrix(NA,V,length(gamma)) 
  for (v in 1:V){
    X.learn = X[folds!=v,]; Y.learn = Y[folds!=v]
    X.val = X[folds==v,]; Y.val = Y[folds==v]
    for (g in 1:length(gamma)){
      f = svm(X.learn,Y.learn,type="C",
              kernel="radial",gamma = gamma[g])
      class = predict(f,X.val)
      err[v,g] = sum(class != Y.val) / length(Y.val)
    }
  }
  plot(gamma,colMeans(err),type='b',log = "x")
  # Return the best gamma
  return(gamma[which.min(colMeans(err))]) 
}

g.star = svm.bestGamma(X,Y)
g.star
```

> Exercise: considering the original iris problem (3 classes, 4 dimensions), use a double CV procedure to compare the appropriate classification methods we studied.

```{r}
library(MASS)
data(iris)
X = iris[,-5]
Y = iris[,5]
```

```{r}
library(class)
knn.bestK <- function(X,Y,V=15,K.max=30){
  folds = rep(1:V,nrow(X)/V)
  err.knn = matrix(NA,V,K.max) 
  for (v in 1:V){
    X.learn = X[folds!=v,]; Y.learn = Y[folds!=v]
    X.val = X[folds==v,]; Y.val = Y[folds==v]
    for (k in 1:K.max){
      class = knn(X.learn,X.val,Y.learn,k=k)
      err.knn[v,k] = sum(class != Y.val) / length(Y.val)
    }
  }
  return(which.min(colMeans(err.knn))) # Return the best K
}
```


```{r}
V = 15
folds = rep(1:V,nrow(X)/V)
err.lda = err.qda = err.knn = rep(NA,V)
err.lin = err.poly = err.rbf = rep(NA,V)
K.star = d.star = g.star = rep(NA,V)
for (v in 1:V){
  X.learn = X[folds!=v,]; Y.learn = Y[folds!=v]
  X.val = X[folds==v,]; Y.val = Y[folds==v]
  # LDA
  f = lda(X.learn,Y.learn)
  class = predict(f,X.val)$class
  err.lda[v] = sum(class != Y.val) / length(Y.val)
  # QDA
  f2 = qda(X.learn,Y.learn)
  class2 = predict(f2,X.val)$class
  err.qda[v] = sum(class2 != Y.val) / length(Y.val)
  # KNN
  K.star[v] = knn.bestK(X.learn,Y.learn)
  class = knn(X.learn,X.val,Y.learn,k=K.star[v])
  err.knn[v] = sum(class != Y.val) / length(Y.val)
  # SVM Linear
  svm.lin = svm(X.learn,Y.learn,type='C',kernel = "linear")
  class = predict(svm.lin,X.val)
  err.lin[v] = sum(class != Y.val) / length(Y.val)
  # SVM poly
  d.star[v] = svm.bestD(X,Y)
  svm.poly = svm(X.learn,Y.learn,type='C',
                 kernel = "poly",degree = d.star[v])
  class = predict(svm.poly,X.val)
  err.poly[v] = sum(class != Y.val) / length(Y.val)
  # SVM RBF
  g.star[v] = svm.bestGamma(X,Y)
  svm.rbf = svm(X.learn,Y.learn,type='C',
                 kernel = "radial",gamma = g.star[v])
  class = predict(svm.rbf,X.val)
  err.rbf[v] = sum(class != Y.val) / length(Y.val)
}
cat('LDA cv. error:',mean(err.lda),'+/-',sd(err.lda),'\n')
cat('QDA cv. error:',mean(err.qda),'+/-',sd(err.qda),'\n')
cat('KNN cv. error:',mean(err.knn),'+/-',sd(err.knn),'\n')
cat('\t-> Best average K:',median(K.star),'\n')
cat('SVM lin cv. error:',mean(err.lin),'+/-',sd(err.lin),'\n')
cat('SVM poly cv. error:',mean(err.poly),'+/-',sd(err.poly),'\n')
cat('\t-> Best average d:',median(d.star),'\n')
cat('SVM RBF cv. error:',mean(err.rbf),'+/-',sd(err.rbf),'\n')
cat('\t-> Best average g:',median(g.star),'\n')
```

# Unsupervised learning: dimension reduction

## PCA

In order to illustrate PCA, let's consider the `swiss` data that are available directly in R:

```{r}
data(swiss)
?swiss
```

For doing a PCA on those data, we can do:

```{r}
X = swiss
pc = princomp(X)
pc
```

```{r}
summary(pc)
```

This display allows you to apply the rule of the 90%: here, with $d=2$ we already have about 93% of the full variance.

If we prefer to select $d$ using the elbow rule on the eigenvalue scree, we can use the following code:

```{r}
screeplot(pc)
```

Here, we will probably select $d=3$ components.

For interpretation, we may rely on the correlation circle that can be obtain with the `biplot` function:

```{r}
biplot(pc)
```
In black are the projection of the data on the two first principle component. In red the correlation circle. We should split them in two plot to be more readable. 
```{r}
par(mfrow=c(1,2))
biplot(pc,col=c(0,2)); box()
biplot(pc,col=c(1,0))
```
from the red, we see that component 1 is mostly related to Catholic variable. 

we can say that the village Oron is protestant that are higly involve in Agriculture.
The village Geneve is almost a mix between catholic and protestant with very low level of aggriculture. 
The Village Herens is higlty Catholic and aggriculture. 
The village Rive GAuche is a mix between catholic and protestant but the level of aggriculture is less than the average. 

> Exercise: Replicate this analysis with the Decathlon data. In particular, I want you to "predict" who won the competition? 

```{r}
deca = read.csv("https://husson.github.io/img/decathlon.csv",sep = ';')
X = deca[,2:11]
row.names(X) = deca[,1]
```

```{r}
pc = princomp(X)
pc
```
```{r}
summary(pc)
```

```{r}
screeplot(pc)
```

```{r}
par(mfrow=c(1,2))
biplot(pc,col=c(0,2)); box()
biplot(pc,col=c(1,0))
```
The compoenent one is negative correlated with 1500m. the higer the value of component one, the lower value of 1500m. Meaning the athelte should arrive at minimum time. 

The component two is nagative correlated with Javelot. Clay and Sebrle are good in Javelot. Kizoglou and Karpov they are not good at Jvelot. 
Barras has large value for component one so he is not good with 1500m 
## Sparce PCA

SPCA allows to perform PCA and ease the interpretation of its results. The "official" implementation of SPCA is proposed in the `elasticnet` package for R:

```{r}
#install.packages("elasticnet")
library(elasticnet)
?spca
```

As we have seen in the course, the goal of SPCA is to sparsify the loading matrix $U$ of PCA:


```{r}
#the first 2 column of PCA transformation matrix U
pc = princomp(X)
d = 2
U = unclass(pc$loadings)
round(U[,1:d],3)
```

Let's now apply SPCA on this Decathlon data:

```{r}
lambda = 1
out = spca(X,d,para = rep(lambda,d))
round(out$loadings,3)
```

At this stage, it is interesting to compare the original PCA, its sparse version with SPCA and (the bad practice) the thresholded PCA.

```{r}
X = as.matrix(X)
par(mfrow = c(1,3))
U = unclass(pc$loadings[,1:d])
plot(X %*% U, main="Original PCA",type="n")
text(X %*% U,labels=rownames(X))

U = out$loadings
U[,2] = -U[,2]
plot(X %*% U, main="Sparse PCA",type="n")
text(X %*% U,labels=rownames(X))

U = unclass(pc$loadings[,1:d])
U[U<0.01] = 0
plot(X %*% U, main="Thresholded PCA",type="n")
text(X %*% U,labels=rownames(X))
```
We see that the geometric space of pca and spca are very similar and the threholding pca is toltally changing the space 

> Exercise: play with the parameter lambda to see after which value the visualisation becomes very different.

```{r}
par(mfrow = c(2,4))
for (lambda in c(0,0.01,0.1,1,2,5,10,100)){
  U = spca(X,d,para = rep(lambda,d))$loadings
  plot(X %*% U, main=paste("Sparse PCA",lambda),type="n")
  text(X %*% U,labels=rownames(X))
}
```

Looking at this visualization, we may conclude that $\lambda = 5$ is the right level of sparsity:

```{r}
U = spca(X,d,para = rep(5,d))$loadings
U
```
## Probabilistic PCA (PPCA) and gsPPCA

First, PPCA is implemented in the `Rdimtools` package:

```{r}
#install.packages("Rdimtools")
library('Rdimtools')
```

> Exercsise: follow the documentation to compare visually PCA and PPCA on the Decathlon data.

```{r}
X = as.matrix(X)
par(mfrow = c(1,2))

d = 2
pc = princomp(X)
U = unclass(pc$loadings[,1:d])
plot(X %*% U, main="Original PCA",type="n")
text(X %*% U,labels=rownames(X))

PPCA <- do.ppca(X, ndim=2)
U = PPCA$proj
U[,1] = -U[,1]
plot(X %*% U, main="Probabilistic PCA",type="n")
text(X %*% U,labels=rownames(X))
```

Finally, let's consider the Globally Sparse PPCA (Mattei et al. 2016):

```{r}
#install.packages("Bessel")
library("Bessel")
source('https://raw.githubusercontent.com/pamattei/GSPPCA/master/GSPPCA.R')
```

The last file provides an implementation of `gsppca`:

```{r}
out = gsppca(X,d=2)
out$loadings
```

> Exercise: compare the results of PCA, SPCA ($\lambda=5$) and gsPPCA in term of visualization.


```{r}
X = as.matrix(X)
par(mfrow = c(1,3))
U = unclass(pc$loadings[,1:d])
plot(X %*% U, main="Original PCA",type="n")
text(X %*% U,labels=rownames(X))

U = spca(X,d,para = rep(5,d))$loadings
U = out$loadings
U[,2] = -U[,2]
plot(X %*% U, main="Sparse PCA",type="n")
text(X %*% U,labels=rownames(X))

U = out$loadings
U[,2] = -U[,2]
plot(X %*% U, main="gsPPCA",type="n")
text(X %*% U,labels=rownames(X))
```

As we can see here, gsPPCA proposes directly the right level of sparsity without the intervention of the user. as the user don't need to choose parameter






